---
title: "Project 2"
author:
  - name: Matthew Jacoby
    url: http://mjacob72.github.io/
    affiliations:
      - id: JHU BSPH MMI
        name: Department of Molecular Microbilogy and Immunology
        url: https://publichealth.jhu.edu/departments/molecular-microbiology-and-immunology  
description: "Exploring temperature and rainfall in Australia"
categories: [project 2, projects]
---

*This project, as the rest of the course, is adapted from the version [Stephanie C. Hicks](https://www.stephaniehicks.com/) designed and maintained in 2021 and 2022. Check the recent changes to this file through the `r paste0("[GitHub history](https://github.com/lcolladotor/jhustatcomputing/commits/main/", basename(dirname(getwd())), "/", basename(getwd()), "/index.qmd)")`.*

# Background

**Due date: October 4th at 11:59pm**

The goal of this assignment is to practice designing and writing functions along with practicing our tidyverse skills that we learned in our previous project. Writing functions involves thinking about how code should be divided up and what the interface/arguments should be. In addition, you need to think about what the (function will return as output.

### To submit your project

Please write up your project using R Markdown and processed with `knitr`. Compile your document as an **HTML file** and submit your HTML file to the dropbox on Courseplus. Please **show all your code** (i.e. make sure to set `echo = TRUE`) for each of the answers to each part.

### Install packages

Before attempting this assignment, you should first install the following packages, if they are not already installed:

```{r}
#| eval: false
install.packages("tidyverse")
install.packages("tidytuesdayR")
```

# Part 1: Fun with functions

In this part, we are going to practice creating functions.

### Part 1A: Exponential transformation

The exponential of a number can be written as an infinite series expansion of the form 
$$
\exp(x) = 1 + x + \frac{x^2}{2!} + \frac{x^3}{3!} + \cdots
$$

Of course, we cannot compute an infinite series by the end of this term and so we must truncate it at a certain point in the series. The truncated sum of terms represents an approximation to the true exponential, but the approximation may be usable.

Write a (function that computes the exponential of a number using the truncated series expansion. The (function should take two arguments:

-   `x`: the number to be exponentiated

-   `k`: the number of terms to be used in the series expansion beyond the constant 1. The value of `k` is always $\geq 1$.

For example, if $k = 1$, then the `Exp` (function should return the number $1 + x$. If $k = 2$, then you should return the number $1 + x + x^2/2!$.

Include at least one example of output using your (function.

::: callout-note
-   You can assume that the input value `x` will always be a *single* number.

-   You can assume that the value `k` will always be an integer $\geq 1$.

-   Do not use the `exp()` (function in R.

-   The `factorial()` (function can be used to compute factorials.
:::

```{r}
# Define a function that approximates exp(x) using a truncated series
test_exp <- function(x, k) {

    # Initialize result with the constant term (1)
    val <- 1

    # Add terms from x^1/1! through x^k/k!
    if (k > 0){
        for (i in seq_len(k)) {

            # Add current series term to running total
            val <- val + ((x^i)/factorial(i))
        }
    }

    # Return approximated exponential value
    return(val)
}

# Test using test_exp
test_exp(2,5)

# Confirmation with exp
exp(2)

# Test using test_exp
test_exp(c(3,5,7),10)

# Confirmation with exp
exp(3)
exp(5)
exp(7)

```

### Part 1B: Sample mean and sample standard deviation

Next, write two functions called `sample_mean()` and `sample_sd()` that takes as input a vector of data of length $N$ and calculates the sample average and sample standard deviation for the set of $N$ observations.

$$
\bar{x} = \frac{1}{N} \sum_{i=1}^n x_i
$$

$$
s = \sqrt{\frac{1}{N-1} \sum_{i=1}^N (x_i - \overline{x})^2}
$$

 As a primer, the $\sum_{i = 1}^3 x_i$ is equivalent to doing the sum of $x_i$ starting from $i = 1$, increasing $i$ by 1 each then, up to when $i = 3$. That is, $x_1 + x_2 + x_3$.

Include at least one example of output using your functions.

::: callout-note
-   You can assume that the input value `x` will always be a *vector* of numbers of length *N*.

-   Do not use the `mean()` and `sd()` functions in R.
:::



```{r}
# Function to calculate the sample mean
sample_mean <- function(x) {

    # Compute total sum of values
    tot_sum <- sum(x)

    # Determine sample size
    big_n <- length(x)

    # Calculate average
    averg <- tot_sum/big_n

    # Return sample mean
    return(averg)
}

# Test using sample_mean
sample_mean(c(2,6,3,4,5,9,1,5,2,4,8))

# Confirmation using mean
mean(c(2,6,3,4,5,9,1,5,2,4,8))

# Test using sample_mean
sample_mean(c(324,334,354,345,341,322,347,362,342))

# Confirmation using mean
mean(c(324,334,354,345,341,322,347,362,342))

# Function to calculate sample standard deviation
sample_sd <- function(inp) {

    # Calculate sample mean using custom function
    smpl_m <- sample_mean(inp)

    # Determine sample size
    big_n <- length(inp)

    # Apply sample standard deviation formula
    st_dev <- sqrt(sum((inp - smpl_m)^2)/(big_n - 1))

    # Return standard deviation
    return(st_dev)
}

# Test using sample_sd
sample_sd(c(2,6,3,4,5,9,1,5,2,4,8))

# Confirmation using sd
sd(c(2,6,3,4,5,9,1,5,2,4,8))

# Test using sample_sd
sample_sd(c(324,334,354,345,341,322,347,362,342))

# Confirmation using sd
sd(c(324,334,354,345,341,322,347,362,342))
```

### Part 1C: Confidence intervals

Next, write a function called `calculate_CI()` that:

1.  There should be two inputs to the `calculate_CI()`. First, it should take as input a vector of data of length $N$. Second, the (function should also have a `conf` ($=1-\alpha$) argument that allows the confidence interval to be adapted for different $\alpha$.

2.  Calculates a confidence interval (CI) (e.g. a 95% CI) for the estimate of the mean in the population. If you are not familiar with confidence intervals, it is an interval that contains the population parameter with probability $1-\alpha$ taking on this form


$$
\bar{x} \pm t_{\alpha/2, N-1} s_{\bar{x}}
= ( \bar{x} - t_{\alpha/2, N-1} s_{\bar{x}}, \bar{x} + t_{\alpha/2, N-1} s_{\bar{x}} )
$$

where $t_{\alpha/2, N-1}$ is the value needed to generate an area of $\alpha / 2$ in each tail of the $t$-distribution with $N-1$ degrees of freedom and $s_{\bar{x}} = \frac{s}{\sqrt{N}}$ is the standard error of the mean. For example, if we pick a 95% confidence interval and $N$=50, then you can calculate $t_{\alpha/2, N-1}$ as

```{r}
alpha <- 1 - 0.95
degrees_freedom <- 50 - 1
t_score <- qt(p = alpha / 2, df = degrees_freedom, lower.tail = FALSE)
```

3.  Returns a named vector of length 2, where the first value is the `lower_bound`, the second value is the `upper_bound`.

```{r}
# Function to calculate a confidence interval for the mean
calculate_CI <- function(inp, conf = 0.95) {

    # Determine alpha level
    alpha <- (1 - conf)

    # Degrees of freedom
    degrees_freedom <- (length(inp) - 1)

    # Critical t value
    t_score <- qt(p = alpha / 2, df = degrees_freedom, lower.tail = FALSE)

    # Margin of error
    adjusted_tscore <- t_score * (sample_sd(inp)/sqrt(length(inp)))

    # Lower confidence limit
    lower <- sample_mean(inp) - adjusted_tscore

    # Upper confidence limit
    upper <- sample_mean(inp) + adjusted_tscore

    # Return confidence interval bounds
    return(c(lower, upper))
}

# Test data vector
fit_test <- c(2,6,3,4,5,9,1,5,2,4,8,7,2,3,6,5,8,1,2,5,3,7,8,4,5,2,3,7,4,9,8,2,4,5,3,5,1,2,7,6,3,4,5,8,6,8,7,4,3,4,9,4,5,6,3,5,2,4)

# Calculate confidence interval
calculate_CI(fit_test, conf = 0.95)

# Confirmation test using linear model
test_attempt <- lm(fit_test ~ 1)

# Built-in confidence interval
confint(test_attempt, level = 0.95)

```

Include example of output from your (function showing the output when using two different levels of `conf`.


::: callout-note
If you want to check if your (function output matches an existing (function in R, consider a vector $x$ of length $N$ and see if the following two code chunks match.

```{r, eval=FALSE}
calculate_CI(x, conf = 0.95)
```

```{r,eval=FALSE}
dat <- data.frame(x = x)
fit <- lm(x ~ 1, dat)

Calculate a 95% confidence interval
confint(fit, level = 0.95)
```
:::

# Part 2: Wrangling data

In this part, we will practice our wrangling skills with the tidyverse that we learned about in module 1.

### Data

The two datasets for this part of the assignment comes from [TidyTuesday](https://www.tidytuesday.com). Specifically, we will use the following data from January 2020, which I have provided for you below:

```{r, eval=FALSE}
#tuesdata <- tidytuesdayR::tt_load("2020-01-07")
#rainfall <- tuesdata$rainfall
#temperature <- tuesdata$temperature
```

However, to avoid re-downloading data, we will check to see if those files already exist using an `if()` statement:

```{r, message=FALSE}
library(here)
if (!file.exists(here("data", "tuesdata_rainfall.RDS"))) {
    tuesdata <- tidytuesdayR::tt_load("2020-01-07")
    rainfall <- tuesdata$rainfall
    temperature <- tuesdata$temperature

    if (!dir.exists(here("data"))) {
        dir.create(here("data"), recursive = TRUE)
    }
    # save the files to RDS objects
    saveRDS(tuesdata$rainfall, file = here("data", "tuesdata_rainfall.RDS"))
    saveRDS(tuesdata$temperature, file = here("data", "tuesdata_temperature.RDS"))
}
```

::: callout-note
The above code will only run if it cannot find the path to the `tuesdata_rainfall.RDS` on your computer. Then, we can just read in these files every time we knit the R Markdown, instead of re-downloading them every time.
:::

Let's load the datasets

```{r, eval=TRUE, message=FALSE}
rainfall <- readRDS(here("data", "tuesdata_rainfall.RDS"))
temperature <- readRDS(here("data", "tuesdata_temperature.RDS"))
```

Now we can look at the data with `glimpse()`

```{r,message=FALSE}
library(tidyverse)

glimpse(rainfall)
glimpse(temperature)
```

If we look at the [TidyTuesday github repo](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020#2020-data) from 2020, we see this dataset contains temperature and rainfall data from Australia.

Here is a data dictionary for what all the column names mean:

-   <https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-07/readme.md#data-dictionary>

### Tasks

Using the `rainfall` and `temperature` data, perform the following steps and create a new data frame called `df`:

1.  Start with `rainfall` dataset and drop any rows with NAs.
2.  Create a new column titled `date` that combines the columns `year`, `month`, `day` into one column separated by "-". (e.g. "2020-01-01"). This column should not be a character, but should be recognized as a date. (**Hint**: check out the `ymd()` (function in `lubridate` R package). You will also want to add a column that just keeps the `year`.
3.  Using the `city_name` column, convert the city names (character strings) to all upper case.
4.  Join this wrangled rainfall dataset with the `temperature` dataset such that it includes only observations that are in both data frames. (**Hint**: there are two keys that you will need to join the two datasets together). (**Hint**: If all has gone well thus far, you should have a dataset with 83,964 rows and 13 columns).

::: callout-note
-   You may need to use functions outside these packages to obtain this result, in particular you may find the functions `drop_na()` from `tidyr` and `str_to_upper()` (function from `stringr` useful.
:::

```{r}
# Load required packages
library(tidyverse)
library(dplyr)
library(here)
library(lubridate)

# Clean rainfall data
temp_df <- rainfall %>%
    drop_na() %>%                         # Remove missing values
    mutate("date" = make_date(year, month, day)) %>%   # Create Date column
    select(!c(day, month)) %>%            # Remove day and month columns
    mutate(city_name = toupper(city_name)) # Convert city names to uppercase

# Join rainfall and temperature datasets
df <- temp_df %>%
    inner_join(
        temperature,
        by = c("city_name", "date"),
        relationship = "many-to-many"
    )

# View resulting data frame
df

```

# Part 3: Data visualization

In this part, we will practice our `ggplot2` plotting skills within the tidyverse starting with our wrangled `df` data from Part 2. For full credit in this part (and for all plots that you make), your plots should include:

1.  An overall title for the plot and a subtitle summarizing key trends that you found. Also include a caption in the figure.
2.  There should be an informative x-axis and y-axis label.

Consider playing around with the `theme()` (function to make the figure shine, including playing with background colors, font, etc.

### Part 3A: Plotting temperature data over time

Use the functions in `ggplot2` package to make a line plot of the max and min temperature (y-axis) over time (x-axis) for each city in our wrangled data from Part 2. You should only consider years 2014 and onwards. For full credit, your plot should include:

1.  For a given city, the min and max temperature should both appear on the plot, but they should be two different colors.
2.  Use a facet function to facet by `city_name` to show all cities in one figure.

```{r}
# Load required libraries
library(tidyverse)
library(ggplot2)
library(dplyr)

# Summarize annual minimum and maximum temperatures
simpl_temp <- df %>%
    filter(year >= 2014) %>%
    group_by(year, city_name) %>%
    summarise(
        min_temp = min(temperature, na.rm = TRUE),
        max_temp = max(temperature, na.rm = TRUE),
        .groups = "drop"
    )

# Convert wide format to long format
temp_long <- simpl_temp %>%
    pivot_longer(
        -c(city_name, year),
        names_to = "min_max",
        values_to = "temperature"
    )

# Plot annual min/max temperatures
ggplot(temp_long, aes(x = year, y = temperature, color = min_max)) +
    geom_point() +
    geom_line() +
    facet_wrap(~ city_name) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = +1.0)
    ) +
    labs(
        title = "Min/Max Temperatures in Australian Cities (2014-2019)"
    )

```

### Part 3B: Plotting rainfall over time

Here we want to explore the distribution of rainfall (log scale) with histograms for a given city (indicated by the `city_name` column) for a given year (indicated by the `year` column) so we can make some exploratory plots of the data.

::: callout-note
You are again using the wrangled data from Part 2.
:::

The following code plots the data from one city (`city_name == "PERTH"`) in a given year (`year == 2000`).

```{r, eval=FALSE}
df %>%
    filter(city_name == "PERTH", year == 2000) %>%
    ggplot(aes(log(rainfall))) +
    geom_histogram()
```

While this code is useful, it only provides us information on one city in one year. We could cut and paste this code to look at other cities/years, but that can be error prone and just plain messy.

The aim here is to **design** and **implement** a function that can be re-used to visualize all of the data in this dataset.

1.  There are 2 aspects that may vary in the dataset: The **city_name** and the **year**. Note that not all combinations of `city_name` and `year` have measurements.

2.  Your function should take as input two arguments **city_name** and **year**.

3.  Given the input from the user, your function should return a **single** histogram for that input. Furthermore, the data should be **readable** on that plot so that it is in fact useful. It should be possible visualize the entire dataset with your function (through repeated calls to your function).

4.  If the user enters an input that does not exist in the dataset, your function should catch that and report an error (via the `stop()` function).

For this section,

1.  Write a short description of how you chose to design your function and why.

2.  Present the code for your function in the R markdown document.

3.  Include at least one example of output from your function.

```{r}
# Function to plot rainfall histogram for a city and year
city_year <- function(cityname, recorded_year) {
    cityname <- toupper(cityname)
    subset_data <- df |>
        dplyr::filter(
            city_name == cityname,
            year == recorded_year, 
            rainfall == rainfall
        )
    # Check whether data exist
    if (nrow(subset_data) == 0) {
        stop(
            paste(
                "No measurements found for city_name =",
                cityname, "and year =", recorded_year), call. = FALSE
                )
    }
    
    # Create histogram
    ggplot(subset_data, aes(log(rainfall)))+
    geom_histogram(fill = "blue", color = "gray", bins = 15)+
    labs(title = paste("Total log(rainfall) in", cityname, "for", recorded_year))
}
```

1. I designed my function to take in a string and year, convert the string to uppercase then filter to only look at the subset of data associated with that year and city. I did this as it isolates the necessary variables needed to perform the analysis. I then checked if the subset is empty, which if it is, then the function stops and calls an error. Assuming everything is met in the above, the function then plots a histogram which allows for the title to be changed based on the city and year put into the function. The histogram displays the distribution such that the individual columns are visible and there is an appropriate amount of columns for the majority of the data pairs available. 

# Positive control test
```{r}
city_year("PERTH", 2000)
city_year("Sydney", 2014)
```

# Error test (eval = FALSE to allow page to render)
```{r, eval = FALSE}
#city_year("PERTH", 2053)
```


# Example output
> city_year("PERTH", 2053) <br>
Error: <br>
! No measurements found for city_name = PERTH and year = 2053 <br>
Show Traceback <br>

# Part 4: Apply functions and plot

### Part 4A: Tasks

In this part, we will apply the functions we wrote in Part 1 to our rainfall data starting with our wrangled `df` data from Part 2.

1.  First, filter for only years including 2014 and onwards.
2.  For a given city and for a given year, calculate the sample mean (using your function `sample_mean()`), the sample standard deviation (using your function `sample_sd()`), and a 95% confidence interval for the average rainfall (using your function `calculate_CI()`). Specifically, you should add two columns in this summarized dataset: a column titled `lower_bound` and a column titled `upper_bound` containing the lower and upper bounds for you CI that you calculated (using your function `calculate_CI()`).
3.  Call this summarized dataset `rain_df`.

```{r}
# Create yearly rainfall summary data frame
rain_df <- df %>%
    filter(year >= 2014) %>%
    group_by(year, city_name) %>%
    summarise(
        avg_rainfall = round(sample_mean(rainfall), digits = 2),
        std_dev = round(sample_sd(rainfall), digits = 2),
        lower_bound = round(calculate_CI(rainfall, conf = 0.95)[1], digits = 2),
        upper_bound = round(calculate_CI(rainfall, conf = 0.95)[2], digits = 2),
        .groups = "drop_last"
    ) %>%
    ungroup()
```

### Part 4B: Tasks

Using the `rain_df`, plots the estimates of mean rainfall and the 95% confidence intervals on the same plot. There should be a separate faceted plot for each city. Think about using `ggplot()` with both `geom_point()` (and `geom_line()` to connect the points) for the means and `geom_errorbar()` for the lower and upper bounds of the confidence interval. Check <https://r-graphics.org/RECIPE-ANNOTATE-ERROR-BAR.html> and/or the official documentation <https://ggplot2.tidyverse.org/reference/geom_linerange.html> for examples of how to use `geom_errorbar()`.

```{r}
# Convert city names to title case for plotting
rain_df <- rain_df %>%
    mutate(city_name = str_to_title(str_to_lower(city_name)))

# Plot mean rainfall with confidence intervals
ggplot(
    rain_df,
    aes(
        x = year,
        y = avg_rainfall,
        ymin = lower_bound,
        ymax = upper_bound,
        color = city_name
    )
) +
    geom_errorbar() +
    geom_point(aes(x = year, y = avg_rainfall)) +
    geom_line() +
    facet_grid(~ city_name) +
    theme(
        axis.text.x = element_text(angle = 45, hjust = +1.0)
    ) +
    labs(
        title = "Average Annual Rainfall in Australian Cities (2014-2019)",
        subtitle = "Decreasing trend in annual rainfall.",
        y = "Average Rainfall (in)"
    ) +
    scale_color_discrete(
        name = "City"
    )
```

# R session information

```{r}
options(width = 120)
sessioninfo::session_info()
```
